---
title: 'DCEG Microbiome Course 2026: NHANES tutorial'
output:
  pdf_document:
    toc: true
    number_sections: true
date: "2026-05-19"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 60)
)

```

## Overview

In this tutorial, we will use R to explore the publicly available [NHANES 2009-2012 Oral Microbiome Data](https://wwwn.cdc.gov/Nchs/Nhanes/omp/).

Before we start, make sure you have a `data` folder containing the following data files (text files downloaded from the NHANES website) to load into R (we will be using the 'Remove SV1032' versions):

-   alpha diversity: `dada2rsv-alpha.txt`
    -   Shannon index
    -   Observed OTUs
    -   Inverse-Simpson index
    -   Faith's Phylogenetic diversity
-   beta diversity
    -   Bray-Curtis: `dada2rsv-braycurtis-beta.txt`
    -   unweighted UniFrac: `dada2rsv-unwunifrac-beta.txt`
    -   weighted UniFrac: `dada2rsv-wunifrac-beta.txt`
-   absolute abundance (read count)
    -   phylum level: `dada2rsv-phylum-count.txt`
    -   genus level: `dada2rsv-genus-count.txt`
-   relative abundance
    -   phylum level: `dada2rsv-phylum-relative.txt`
    -   genus level: `dada2rsv-genus-relative.txt`
-   taxonomy name (for absolute/relative abundance data): `dada2rsv-taxonomy-annotate.txt`

## Setting up

First, install and load packages we will be using. If you already have the package installed, skip the installation and just load.

```{r message=FALSE, warning=FALSE}
## install packages if needed (only need to do this once!)
# install.packages("tidyverse")
# install.packages("haven")
# install.packages("dplyr")
# install.packages("survey")
#tinytex::install_tinytex()


# load packages
library(haven)
library(survey)
library(plyr)
library(dplyr)
library(tidyverse)
library(readr)
library(tinytex)
```

Now check your working directory. This should be the parent folder that contains your `data` folder.\
If you need to change your working directory, use `setwd()` and insert the file path you would like to change to.

```{r results='hide'}
getwd()

```

## Import raw data

Note: We will only be using the Bray-Curtis beta diversity matrix for this tutorial, but you can look at the other types on your own.

```{r results='hide'}
# alpha diversity
raw_data_alpha <- read_delim("dada2rsv-alpha.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE, na=".")

# beta diversity
raw_data_beta_bc <- read_delim("dada2rsv-braycurtis-beta.txt",  
    delim = "\t", escape_double = FALSE,
    trim_ws = TRUE, na=".")

# absolute abundance
raw_data_aa <- read_delim("dada2rsv-genus-count.txt",
    delim = "\t", escape_double = FALSE,
    trim_ws = TRUE)

# relative abundance
raw_data_ra <- read_delim("dada2rsv-genus-relative.txt",
    delim = "\t", escape_double = FALSE,
    trim_ws = TRUE)

# taxa names 
raw_data_taxaname <- read_delim("dada2rsv-taxonomy-annotate.txt",
    delim = "\t", escape_double = FALSE,
    trim_ws = TRUE)

# demographic data
raw_data_demo_f <- read_xpt("DEMO_F.xpt") #2009-2010
raw_data_demo_g <- read_xpt("DEMO_G.xpt") #2011-2012

# diet data
raw_data_diet_f <- read_xpt("DR1TOT_F.xpt") #2009-2010
raw_data_diet_g <- read_xpt("DR1TOT_G.xpt") #2011-2012
```

Take a peek at the data...

**Alpha diversity:** Here, `SEQN` is the participant ID (rows), and the columns represent values for four different alpha diversity metrics (i.e., shannon, observed ASVs, Faith's PD, Shannon index, and Inverse Simpson's index) at different sequencing depths, over 10 iterations of resampling.

```{r}
tibble(raw_data_alpha)
```

**Beta diversity (Bray-Curtis):** This is a matrix of samples by samples.

```{r}
tibble(raw_data_beta_bc)
```

**Absolute/relative abundance:** Samples by taxa, with counts for absolute abundance and relative abundance calculated by dividing taxon counts by total counts per sample. Taxa are represented by index numbers - we will use `data_taxaname` later to link names.

```{r}
tibble(raw_data_aa)
tibble(raw_data_ra)
```

**Demographic/diet data:**
```{r}
tibble(raw_data_demo_f)
tibble(raw_data_demo_g)

tibble(raw_data_diet_f)
tibble(raw_data_diet_g)
```

## Set up analytic dataset

### clean covariates

We will merge the demographic and diet data, and clean some of the variables we will be using. 

```{r}
# Append the two cycles of demographic and diet data
data_demo <- bind_rows(raw_data_demo_f, raw_data_demo_g) #demographic data
data_diet <- bind_rows(raw_data_diet_f, raw_data_diet_g) #diet data

dim(data_demo)
dim(data_diet)

# Merge the demopgraic and diet data into one covariate data
data_cov <- full_join(data_demo, data_diet, by = "SEQN")
dim(data_cov)
```

Create a new survey weight variable: since we are combining 2 NHANES cycles (2009-2010, 2011-2012), we need to create weights for the 4 years:

```{r}
data_cov$WTMEC4YR <- data_cov$WTMEC2YR/2
summary(data_cov$WTMEC2YR)
summary(data_cov$WTMEC4YR)
```

Now we will clean the sex, race/ethnicity, and age variables. 

```{r}
# sex (RIAGENDR)
data_cov$RIAGENDR <- factor(data_cov$RIAGENDR, 
                            levels = c(1, 2), labels = c("Male", "Female"))

# race/ethnicity (RIDRETH1)
data_cov$RIDRETH1 <- factor(data_cov$RIDRETH1, levels = c(1, 2, 3, 4, 5), 
                            labels = c("Mexican American", "Other Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Other"))

#age (RIDAGEYR)
data_cov <- data_cov %>%
  mutate(RIDAGEYR_cat = case_when(
    RIDAGEYR < 20 ~ "0-19",
    RIDAGEYR >= 20 & RIDAGEYR < 40 ~ "20-39",
    RIDAGEYR >= 40 & RIDAGEYR < 60 ~ "40-59",
    RIDAGEYR >= 60 ~ "60+",
    TRUE ~ NA_character_
  )) %>% 
  mutate(
    RIDAGEYR_cat = factor(RIDAGEYR_cat, levels = c("0-19", "20-39", "40-59", "60+"))
  )
```

Use the weights to look at counts and proportions of sex, race/ethnicity, and age.

```{r}
# set up survey design
nhanes_design <- svydesign(
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC4YR,
  data = data_cov,
  nest = TRUE
)

# sex 
table(data_cov$RIAGENDR, useNA = "ifany") # unweighted
svytable(~RIAGENDR, nhanes_design) # weighted

# race/ethnicity
table(data_cov$RIDRETH1, useNA = "ifany") # unweighted
svytable(~RIDRETH1, nhanes_design) # weighted

# age
table(data_cov$RIDAGEYR_cat, useNA = "ifany") # unweighted
svytable(~RIDAGEYR_cat, nhanes_design) # weighted
```

We will also look at differences in the microbiome by quartile of caffeine intake - create a variable for caffeine quartiles

```{r}
# caffeine (DR1TCAFF)
quartiles <- coef(
  svyquantile(~DR1TCAFF, nhanes_design,
              quantiles = c(0.25, 0.5, 0.75),
              na.rm = TRUE))
print(quartiles)
q_cut <- as.numeric(quartiles)

data_cov$quart_DR1TCAFF <- cut(data_cov$DR1TCAFF, # create quartiles
                                    breaks = c(-Inf, q_cut, Inf),
                                    labels = c("Q1","Q2","Q3","Q4"),
                                    include.lowest = TRUE)

nhanes_design <- update(nhanes_design, quart_DR1TCAFF = data_cov$quart_DR1TCAFF) #update survey design

# age
table(data_cov$quart_DR1TCAFF, useNA = "ifany") # unweighted
svytable(~quart_DR1TCAFF, nhanes_design) # weighted counts
svymean(~factor(quart_DR1TCAFF), nhanes_design, na.rm=TRUE)  # weighted proportions
```

### subset data

We will subset the covariate and microbiome data to include the same participants in each dataset.

```{r}
# find SEQN ids included in all datasets
seqn_sub <- Reduce(intersect, list(
  raw_data_alpha$SEQN,
  raw_data_beta_bc$SEQN,
  raw_data_aa$SEQN,
  raw_data_ra$SEQN
))

length(seqn_sub)
sum(duplicated(seqn_sub)) # no duplicates
```

We will include these `r length(seqn_sub)` participants in the subsequent analyses.


## Alpha diversity analyses

Merge the covariate data with the alpha diversity data.

```{r}
# alpha diversity
data_alpha <- raw_data_alpha %>% 
  left_join(data_cov, by="SEQN")

dim(data_alpha)
```

Using the rarefaction of 10,000, we will first average together the 10 resamples for each alpha diversity metric.

```{r out.width="50%"}
# Shannon index
data_alpha$shannon <- rowMeans(data_alpha[, c(
    "RSV_ShanWienDiv_10000_0", "RSV_ShanWienDiv_10000_1", "RSV_ShanWienDiv_10000_2", "RSV_ShanWienDiv_10000_3",
    "RSV_ShanWienDiv_10000_4", "RSV_ShanWienDiv_10000_5", "RSV_ShanWienDiv_10000_6", "RSV_ShanWienDiv_10000_7",
    "RSV_ShanWienDiv_10000_8", "RSV_ShanWienDiv_10000_9")], na.rm = TRUE)

# Observed OTUs
data_alpha$observedotus <- rowMeans(data_alpha[, c(
    "RSV_ObservedOTUs_10000_0", "RSV_ObservedOTUs_10000_1", "RSV_ObservedOTUs_10000_2", "RSV_ObservedOTUs_10000_3",
    "RSV_ObservedOTUs_10000_4", "RSV_ObservedOTUs_10000_5", "RSV_ObservedOTUs_10000_6", "RSV_ObservedOTUs_10000_7",
    "RSV_ObservedOTUs_10000_8", "RSV_ObservedOTUs_10000_9")], na.rm = TRUE)

# Faith's PD
data_alpha$faithpd <- rowMeans(data_alpha[, c(
    "RSV_FaPhyloDiv_10000_0", "RSV_FaPhyloDiv_10000_1", "RSV_FaPhyloDiv_10000_2", "RSV_FaPhyloDiv_10000_3",
    "RSV_FaPhyloDiv_10000_4", "RSV_FaPhyloDiv_10000_5", "RSV_FaPhyloDiv_10000_6", "RSV_FaPhyloDiv_10000_7",
    "RSV_FaPhyloDiv_10000_8", "RSV_FaPhyloDiv_10000_9")], na.rm = TRUE)

# Inverse Simpson
data_alpha$invsimp <- rowMeans(data_alpha[, c(
    "RSV_InverseSimpson_10000_0", "RSV_InverseSimpson_10000_1", "RSV_InverseSimpson_10000_2", "RSV_InverseSimpson_10000_3",
    "RSV_InverseSimpson_10000_4", "RSV_InverseSimpson_10000_5", "RSV_InverseSimpson_10000_6", "RSV_InverseSimpson_10000_7",
    "RSV_InverseSimpson_10000_8", "RSV_InverseSimpson_10000_9")], na.rm = TRUE)

#Examine the distribution for each alpha diversity metric -- this is unweighted; WEIGHTED VERSION BELOW
hist(data_alpha$shannon)
hist(data_alpha$observedotus) #Note: these are actually ASVs, the dataset just named them OTUs (a more dated method, ASVs is more often used now) 
hist(data_alpha$invsimp)
hist(data_alpha$faithpd)

#Weighted versions of the distribution of each alpha diversity metric
ggplot(data_alpha, aes(x = shannon, weight = WTMEC4YR)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(title = "Weighted histogram of Shannon index", x = "Shannon", y = "Weighted count") +
  theme_minimal()

ggplot(data_alpha, aes(x = observedotus, weight = WTMEC4YR)) + 
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(title = "Weighted histogram of ASVs", x = "observed ASVs", y = "Weighted count") +
  theme_minimal()

ggplot(data_alpha, aes(x = invsimp, weight = WTMEC4YR)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(title = "Weighted histogram of Inverse-simpson index", x = "Inverse Simpson", y = "Weighted count") +
  theme_minimal()

ggplot(data_alpha, aes(x = faithpd, weight = WTMEC4YR)) +
  geom_histogram(bins = 30, na.rm = TRUE) +
  labs(title = "Weighted histogram of Faith's Phylogenetic Diversity", x = "Faith's PD", y = "Weighted count") +
  theme_minimal()

#recreate the survey design object now that all variables have been created:
nhanes_design_alpha <- svydesign(
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC4YR,
  data = data_alpha,
  nest = TRUE
)

# subset
nhanes_design_alpha <- subset(nhanes_design_alpha, SEQN %in% seqn_sub)
nrow(nhanes_design_alpha) # 9349

#Running regression models to see what relationships exist between the alpha diversity measures and sex, age category, race/ethnicity, and caffeine intake (quartiles) (note: other variables would need to be adjusted for to have a more reliable model - you can add in other variables on your own)

#Model 1: Sex = independent variable, alpha-div = dependent variable (adjusted for age, race/ethnicity)
#note: you may get a warning message after this which occurs because some observations in the survey design have weight = 0 and the svyglm() function will ignore them when estimating the dispersion parameter
model_1_shannon <- svyglm(shannon ~ RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_1_shannon)

model_1_observedotus <- svyglm(observedotus ~ RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_1_observedotus)

model_1_invsimp <- svyglm(invsimp ~ RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_1_invsimp)

model_1_faithpd <- svyglm(faithpd ~ RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_1_faithpd)


#Model 2: Caffeine (quartiles) = independent variable, alpha-div = dependent variable (adjusted for sex, age, race/ethnicity)

#Note: the alpha div measures have 313 missing values; ggplot() automatically drops them and will give you a warning, but it's fine - nothing is wrong with your code
#Shannon
ggplot(data_alpha, aes(x = shannon, weight = WTMEC4YR)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~quart_DR1TCAFF) +
  labs(
    title = "Distribution of Shannon Diversity by Caffeine Quartile",
    x = "Shannon Diversity Index",
    y = "Count"
  ) +
  theme_minimal()

#Observed OTUs
ggplot(data_alpha, aes(x = observedotus, weight = WTMEC4YR)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~quart_DR1TCAFF) +
  labs(
    title = "Distribution of Observed OTUs by Caffeine Quartile",
    x = "Observed OTUs",
    y = "Count"
  ) +
  theme_minimal()

#Inverse-Simpson
ggplot(data_alpha, aes(x = invsimp, weight = WTMEC4YR)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~quart_DR1TCAFF) +
  labs(
    title = "Distribution of Inverse-Simpson Index by Caffeine Quartile",
    x = "Inverse-Simpson Index",
    y = "Count"
  ) +
  theme_minimal()

#Faith's PD
ggplot(data_alpha, aes(x = faithpd, weight = WTMEC4YR)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  facet_wrap(~quart_DR1TCAFF) +
  labs(
    title = "Distribution of Faith's Phylogenetic Diversity by Caffeine Quartile",
    x = "Faiths PD",
    y = "Count"
  ) +
  theme_minimal()

# boxplot to compare across caffeine quartiles
ggplot(data_alpha, aes(x = quart_DR1TCAFF, y = shannon, weight = WTMEC4YR, color = quart_DR1TCAFF)) +
  geom_point(alpha = 0.05, position=position_jitterdodge(dodge.width=1)) +
  geom_boxplot(width=0.5, outlier.shape = NA, fill=NA, linewidth=1) +
  theme_minimal()

ggplot(data_alpha, aes(x = quart_DR1TCAFF, y = observedotus, weight = WTMEC4YR, color = quart_DR1TCAFF)) +
  geom_point(alpha = 0.05, position=position_jitterdodge(dodge.width=1)) +
  geom_boxplot(width=0.5, outlier.shape = NA, fill=NA, linewidth=1) +
  theme_minimal()

ggplot(data_alpha, aes(x = quart_DR1TCAFF, y = invsimp, weight = WTMEC4YR, color = quart_DR1TCAFF)) +
  geom_point(alpha = 0.05, position=position_jitterdodge(dodge.width=1)) +
  geom_boxplot(width=0.5, outlier.shape = NA, fill=NA, linewidth=1) +
  theme_minimal()

ggplot(data_alpha, aes(x = quart_DR1TCAFF, y = faithpd, weight = WTMEC4YR, color = quart_DR1TCAFF)) +
  geom_point(alpha = 0.05, position=position_jitterdodge(dodge.width=1)) +
  geom_boxplot(width=0.5, outlier.shape = NA, fill=NA, linewidth=1) +
  theme_minimal()


#running the regression models to see if there are statistical differences in each alpha diversity metric by quartile of caffeine intake, adjusting for sex and race/ethnicity
model_2_shannon <- svyglm(shannon ~ quart_DR1TCAFF + RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_2_shannon)

model_2_observedotus <- svyglm(observedotus ~ quart_DR1TCAFF + RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_2_observedotus)

model_2_invsimp <- svyglm(invsimp ~ quart_DR1TCAFF + RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_2_invsimp)

model_2_faithpd <- svyglm(faithpd ~ quart_DR1TCAFF + RIAGENDR + RIDAGEYR_cat + RIDRETH1, design = nhanes_design_alpha)
summary(model_2_faithpd)
```

## Beta diversity analyses

```{r}

data_beta_bc <- as.data.frame(raw_data_beta_bc)

#Check Bray-Curtis matrix dimensions
dim(data_beta_bc) #9349x9350

#Store first column (SEQN) as row names to create square matrix
rownames(data_beta_bc) <- data_beta_bc$SEQN
data_beta_bc$SEQN <- NULL

#Recheck dimensions
dim(data_beta_bc)  #9349x9349

#Check that row names match column names
all(rownames(data_beta_bc) == colnames(data_beta_bc))

#Subset samples to save time for the ordination
data_beta_bc_subset <- data_beta_bc[1:1000, 1:1000]

#Subset NHANES object to match number of observations in data_beta_bc 
nhanes_design_subset <- subset(nhanes_design, SEQN %in% rownames(data_beta_bc_subset))

#Check that subset matches bray-curtis data
length(nhanes_design_subset$variables$SEQN)
length(rownames(data_beta_bc_subset))
setequal(nhanes_design_subset$variables$SEQN, rownames(data_beta_bc_subset))

######################
###Create PCoA Plots##
######################

#Install and load packages
#install.packages("vegan")
library(vegan)

#calculate the first 3 PCoA Axes
pcoa_weight <- wcmdscale(data_beta_bc_subset, 3, eig = TRUE, w=nhanes_design_subset$variables$WTMEC4YR) 

#PCOA PLOT 1: Basic plot
pcoa_plot <- data.frame(
  PCoA1 = pcoa_weight$points[,1],
  PCoA2 = pcoa_weight$points[,2]
)

ggplot(pcoa_plot, aes(x=PCoA1, y=PCoA2)) +
  geom_point() +
  theme_classic() 

table(nhanes_design_subset$variables$quart_DR1TCAFF)

#PCOA PLOT 2: Fill color by caffeine quartile (or any other variable)
pcoa_plot2 <- data.frame(
  PCoA1 = pcoa_weight$points[,1],
  PCoA2 = pcoa_weight$points[,2],
  Caffiene = factor(nhanes_design_subset$variables$quart_DR1TCAFF)
)

pcoa_plot2 %>% 
  filter(!is.na(Caffiene)) %>% #remove those with missing caffeine values
  ggplot(aes(x=PCoA1, y=PCoA2, color=Caffiene)) +
    geom_point() +
    scale_color_manual(values = c("#5a7d4f", "#9e9bff", "#13c34d", "#c99200"),  #color palette options: https://medialab.github.io/iwanthue/
                     labels = c("Q1", "Q2", "Q3", "Q4")) +
    theme_classic() 


##PCOA PLOT 3: Calculate % variance explained by axes and add to axis labels
var_percent <- 100*(pcoa_weight$eig/sum(pcoa_weight$eig[pcoa_weight$eig>0]))  

#Add % variance to axis labels, rounding to two decimal places
xlab <- paste0("PCoA Axis 1 (", round(var_percent[1],2), "%)")
ylab <- paste0("PCoA Axis 2 (", round(var_percent[2],2), "%)")

#Plot
pcoa_plot2 %>% 
  filter(!is.na(Caffiene)) %>% #remove those with missing caffeine values
  ggplot(aes(x=PCoA1, y=PCoA2, color=Caffiene)) +
    geom_point() +
    scale_color_manual(values = c("#5a7d4f", "#9e9bff", "#13c34d", "#c99200"),  #color palette options: https://medialab.github.io/iwanthue/
                     labels = c("Q1", "Q2", "Q3", "Q4")) +
    xlab(xlab) +
    ylab(ylab) +
    theme_classic() 


## You can additionally test for differences between groups as well as associations between overall beta diversity and covariates using PERMANOVA and MiRKAT


```

## Abundance analyses (relative abundance and presence/absence)

### presence/absence - prevalence

We will use the absolute abundance data (counts) to create a presence/absence data (binary - 0s and 1s) and use this to calculate the prevalence of each genus.

```{r}
data_aa <- raw_data_aa 

dim(data_aa) # 9848x1350
sum(duplicated(data_aa$SEQN)) # no duplicates
sum(is.na(data_aa$SEQN)) # 1 missing id

data_aa %>% 
  filter(is.na(SEQN)) %>% 
  head()

# summary statistics of entire abs abundance data
summary(as.vector(as.matrix(data_aa %>% select(-SEQN))))


####
# convert abs abundance to presence/absence data 

data_pa <- data_aa %>% 
  filter(!is.na(SEQN)) %>% 
  mutate_at(vars(-SEQN), function(x){
    x = ifelse(x>0, 1, x)
  })
  
dim(data_pa)
table(as.vector(as.matrix(data_pa %>% select(-SEQN)))) # these are all 0s and 1s now


####
# add diet/demo data 

dim(data_pa)
data_pa <- full_join(data_cov, data_pa, by="SEQN")
dim(data_pa)


####
# design object

nhanes_design_pa <- svydesign(
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC4YR,
  data = data_pa,
  nest = TRUE
)

nhanes_design_pa <- subset(nhanes_design_pa, SEQN %in% seqn_sub)
nrow(nhanes_design_pa) # 9349


####
# calculate prevalence (=mean)

# unweighted
taxa_sumstat <- data_pa %>% 
  filter(SEQN %in% seqn_sub) %>% 
  summarise_at(vars(c(colnames(data_pa)[grepl("RSV_", colnames(data_pa))])), function(x){
    sum(x)/length(x)
  }) %>% 
  pivot_longer(everything(.), names_to = "taxa_id", values_to = "unw_prev")

# weighted

taxa_id <- colnames(data_pa)[grepl("RSV_", colnames(data_pa))]
length(taxa_id)

temp_pa <- list()

for(i in taxa_id){
  temp_pa[[i]] <- svymean(as.formula(paste0("~",i)), nhanes_design_pa, na.rm = TRUE)
}

temp_pa <- llply(temp_pa, as.data.frame) %>% 
  llply(., function(x){
  x %>% 
    rename_with(~c("prev", "prev_se"), .cols=c(1,2))
  }) %>% 
  ldply(.id = "taxa_id") %>% 
  mutate(taxa_id = as.character(taxa_id))

head(temp_pa)

# merge with unweighted prev
dim(taxa_sumstat) # 1349x2
dim(temp_pa) # 1349x3

taxa_sumstat <- taxa_sumstat %>% 
  full_join(., temp_pa, by="taxa_id")

dim(taxa_sumstat)
head(taxa_sumstat)

summary(taxa_sumstat$unw_prev) # unweighted prevalence
summary(taxa_sumstat$prev) # weighted prevalence

rm(temp_pa)
```

### mean relative abundance

We will use the relative abundance data to calculate the mean relative abundance for each genus.

```{r}
data_ra <- raw_data_ra

dim(data_ra) # 9848

sum(is.na(data_ra$SEQN)) # one
sum(duplicated(data_ra$SEQN)) # none

data_ra <- data_ra %>% 
  filter(!is.na(SEQN))

####
# add diet/demo data 

dim(data_ra)
data_ra <- full_join(data_cov, data_ra, by="SEQN")
dim(data_ra)

# check all subject IDs are the same 
identical(data_pa$SEQN, data_ra$SEQN) # true

# check taxa_ids are the same 
identical(colnames(data_pa)[grepl("RSV_", colnames(data_pa))],
          colnames(data_ra)[grepl("RSV_", colnames(data_ra))]) # true


####
# design object

nhanes_design_ra <- svydesign(
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC4YR,
  data = data_ra,
  nest = TRUE
)

nhanes_design_ra <- subset(nhanes_design_ra, SEQN %in% seqn_sub)
nrow(nhanes_design_ra) # 9349


####
# calculate mean relative abundance

# unweighted 
temp_ra_unw <- data_ra %>% 
  filter(SEQN %in% seqn_sub) %>% 
  summarise_at(vars(c(colnames(data_ra)[grepl("RSV_", colnames(data_ra))])),
               mean, na.rm = TRUE) %>% 
  pivot_longer(everything(.), names_to = "taxa_id", values_to = "unw_meanra")


# weighted

identical(colnames(data_ra)[grepl("RSV_", colnames(data_ra))], taxa_id) # true

temp_ra <- list()

for(i in taxa_id){
  temp_ra[[i]] <- svymean(as.formula(paste0("~",i)), nhanes_design_ra, na.rm = TRUE)
}

temp_ra <- llply(temp_ra, as.data.frame) %>% 
  llply(., function(x){
  x %>% 
    rename_with(~c("meanra", "meanra_se"), .cols=c(1,2))
  }) %>% 
  ldply(.id = "taxa_id") %>% 
  mutate(taxa_id = as.character(taxa_id))

head(temp_ra)

dim(taxa_sumstat) # 1349x4
dim(temp_ra_unw) # 1349x2
dim(temp_ra) # 1349x3

taxa_sumstat <- taxa_sumstat %>% 
  full_join(., temp_ra_unw, by="taxa_id") %>% 
  full_join(., temp_ra, by="taxa_id")

dim(taxa_sumstat)
head(taxa_sumstat)

summary(taxa_sumstat$unw_meanra) # unweighted mean relative abundance
summary(taxa_sumstat$meanra) # weighted mean relative abundance

rm(temp_ra, temp_ra_unw, i)
```

```{r}
# link taxa name to the summary statistic table
dim(raw_data_taxaname)
head(raw_data_taxaname) # this has the names for all taxnomic levels, from phylum to genus

data_taxaname <- raw_data_taxaname %>% 
  rename(taxa_id = Name, taxa_name = `Taxonomy in SILVA v123`)

taxa_sumstat <- taxa_sumstat %>% 
  left_join(., data_taxaname, by = "taxa_id") %>% 
  select(taxa_id, taxa_name, everything(.))

head(taxa_sumstat)
```

### taxonomic profile (stacked barplot)

We will create a stacked barplot to visualize the microbiome composition by caffeine quartile.

```{r}
# calculate weighted mean relative abundance by caffeine quartile -- this will take a min to run...

temp <- list()

for(i in taxa_id){
  temp[[i]] <- svyby(as.formula(paste0("~",i)), ~quart_DR1TCAFF, nhanes_design_ra, svymean, na.rm = TRUE)
}

temp <- llply(temp, as.data.frame) %>% 
  llply(., function(x){
  x %>% 
    rename_with(~c("meanra", "meanra_se"), .cols=c(2, 3))
  }) %>% 
  ldply(.id = "taxa_id") %>% 
  mutate(taxa_id = as.character(taxa_id))

head(temp)


# add taxa names
temp <- temp %>% 
  left_join(., data_taxaname, by="taxa_id") %>% 
  select(taxa_id, taxa_name, everything(.))

# add the overall mean relative abundance
temp2 <- taxa_sumstat %>% 
  select(taxa_id, taxa_name, meanra, meanra_se) %>% 
  mutate(quart_DR1TCAFF = "Overall")

temp <- rbind(temp, temp2)

rm(temp2)

# label taxa with overall meanra > 0.01
temp_taxa <- temp %>% 
  filter(quart_DR1TCAFF == "Overall" & meanra>0.01) %>% 
  pull(taxa_id)

length(temp_taxa) # 14

temp <- temp %>% 
  mutate(
    taxa_label = case_when(
      taxa_id %in% temp_taxa ~ str_extract(taxa_name, "[^;]+$"),
      TRUE ~ "Other"
    )
  )

# order the taxa by mean relative abundance
taxa_order <- temp %>% 
  filter(quart_DR1TCAFF == "Overall") %>% 
  arrange(desc(meanra)) %>% 
  pull(taxa_label) %>% 
  unique(.)

temp <- temp %>% 
  mutate(
    taxa_label = factor(taxa_label, levels = taxa_order)
  )

temp %>% count(taxa_label)

mycolors <- as.vector(ggsci::pal_ucscgb()(26))
    
ggplot(temp, aes(x = quart_DR1TCAFF, y=meanra, fill=taxa_label)) +
  geom_bar(stat="identity") +
  labs(x=NULL,y="Relative abundance", fill="Genus") +
  scale_fill_manual(values=mycolors[1:length(levels(temp$taxa_label))]) +
  guides(fill=guide_legend(ncol=1)) +
  theme(legend.position = "right", legend.text=element_text(size=8))
```

```{r}
rm(temp, temp_taxa, taxa_order, i)
```

### filter low prevalence taxa

We will now run some regression models. For the models, we will exclude taxa with weighted prevalence less than 1% or greater than 99%, and also taxa with mean relative abundance less than 0.0001

```{r}
# taxa with all 0s
taxa_sumstat %>% 
  filter(prev == 0 & meanra == 0) %>% 
  nrow() # 89 taxa

# taxa with prev <1%
taxa_sumstat %>% 
  filter(prev < 0.01 | prev > 0.99) %>% 
  nrow() # 1131

# taxa with meanra <0.0001
taxa_sumstat %>% 
  filter(meanra < 0.0001) %>% 
  nrow() # 1250

# create a flag for taxa to filter out
taxa_sumstat <- taxa_sumstat %>% 
  mutate(
    filt_flg = case_when(
      prev <0.01 | prev>0.99 ~ 1,
      meanra < 0.0001 ~ 1,
      TRUE ~ 0
    )
  )

taxa_sumstat %>% 
  count(filt_flg) # 90 taxa to keep

# taxa_id of 90 taxa to keep
taxa_id_sub <- taxa_sumstat %>% 
  filter(filt_flg == 0) %>% 
  pull(taxa_id)
length(taxa_id_sub) # 90
```

### pa model

Run the following model: presence/absence of taxon ~ caffeine quartile + age + sex + race/ethnicity
We will run the model separately for each taxon, with multiple testing correction.

```{r}
model_pa <- list()

for(i in taxa_id_sub){
  model_pa[[i]] <- svyglm(as.formula(paste0(i, "~ quart_DR1TCAFF + RIDAGEYR_cat + RIAGENDR + RIDRETH1")), nhanes_design_pa, family=quasibinomial())
}

model_pa_wald <- llply(model_pa, function(x){
  regTermTest(x, "quart_DR1TCAFF")
})

# example result for RSV_genus1007
summary(model_pa[[1]])
model_pa_wald[1]

# make a results table with the wald p-values
model_pa_res <- llply(model_pa_wald, function(x){
  x$p
}) %>% 
  ldply(., .id = "taxa_id") %>% 
  rename(pval = "1")

# multiple testing correction
dim(model_pa_res)
summary(model_pa_res$pval)

model_pa_res <- model_pa_res %>% 
  mutate(
    adj_pval = p.adjust(pval, method = "bonferroni", n=length(pval))
  )

summary(model_pa_res$adj_pval)

# number of significant taxa (after bonferroni)
model_pa_res %>% 
  filter(adj_pval < 0.05)
## there are three

# link taxa name to results table
model_pa_res <- model_pa_res %>% 
  arrange(adj_pval, pval) %>% 
  left_join(., data_taxaname, by = "taxa_id") %>% 
  select(taxa_id, taxa_name, everything(.))

# look at the significant taxa
model_pa_res %>% 
  filter(adj_pval <0.05)

# look at top 10
head(model_pa_res, 10)
```

```{r}
rm(model_pa) # this is a large object - remove to save memory
```


### clr model

Relative abundance data are compositional (data are constrained to a sum of 100%), but many standard statistical methods assume independent, unconstrained data. We will use the centered log-ratio (CLR) transformation to make the data appropriate for standard methods, including regression. CLR expresses the log of each taxon abundance relative to the geometric mean of all taxa in the sample.

```{r}
head(data_aa)

temp <- data_aa %>% 
  filter(!is.na(SEQN)) %>% 
  column_to_rownames(var = "SEQN") %>% 
  as.matrix(.)

summary(as.vector(temp))
sum(temp == 0) # number of 0s

# log and add pseudocount
temp <- as.data.frame(log(temp + 0.001)) %>% 
  rownames_to_column(var="SEQN") %>% 
  arrange(SEQN)

# calculate the geometric mean of the read counts of all taxa within a sample - this is the denominator for the clr transformation
temp_denom <- temp %>% 
  pivot_longer(-SEQN) %>% 
  group_by(SEQN) %>% 
  summarise(samp_geom_mean = mean(value)) %>% 
  ungroup(.) %>% 
  arrange(SEQN) %>% 
  column_to_rownames(var="SEQN") %>% 
  as.matrix(.)

head(temp_denom)

# clr numerator - this is the logged count data (with pseudocount added)
temp_num <- temp %>% 
  column_to_rownames(var="SEQN") %>% 
  as.matrix(.)

dim(temp_denom) # 9847x1 - geometric mean of all taxa in each sample
dim(temp_num) # 9847x1349 - logged with pseudocount

# check sample ids match
identical(rownames(temp_num), rownames(temp_denom)) #true

# clr transform
data_clr <- sweep(temp_num, 1, temp_denom, "-")

data_clr<- as.data.frame(data_clr) %>% 
  rownames_to_column(var="SEQN")

rm(temp, temp_denom, temp_num)

tibble(data_clr)
```

**OPTIONAL**: Instead of doing the CLR transformation manually, you can also use the `microbiome` and `phyloseq` packages to do this.

```{r}
# this section is optional so the code is commented out

## install microbiome and phyloseq packages from Bioconductor
#  if (!require("BiocManager", quietly = TRUE))
#      install.packages("BiocManager")
#  BiocManager::install("phyloseq")
#  BiocManager::install("microbiome")

# library(phyloseq)
# 
# # format data for phyloseq
# temp <- data_aa %>%
#   filter(!is.na(SEQN)) %>%
#   select(-SEQN) %>%
#   t(.)
# 
# colnames(temp) <- data_aa %>%
#   filter(!is.na(SEQN)) %>%
#   pull(SEQN)
# 
# # add pseudocount
# temp <- temp + 0.001
# 
# ## transform data to phyloseq object
# temp_phylo <- otu_table(temp, taxa_are_rows = TRUE)
# 
# ## and put them in one object
# temp_phylo <- phyloseq(temp_phylo)
# str(temp_phylo)
# 
# temp_phylo_clr <- microbiome::transform(temp_phylo, "clr")
# 
# temp_phylo_clr <- as.data.frame(t(otu_table(temp_phylo_clr)))
# 
# temp_phylo_clr <- temp_phylo_clr %>%
#   rownames_to_column(var="SEQN") %>%
#   arrange(SEQN)
# 
# # is this identical to the manually calculated CLR transformed data?
# identical(temp_phylo_clr, data_clr) # true
# 
# tibble(data_clr)
# tibble(temp_phylo_clr)
# 
# rm(temp, temp_phylo, temp_phylo_clr)
```

Add covariates to the CLR transformed data and subset.

```{r}
# add diet/demo data 

dim(data_clr)
data_clr <- data_clr %>% 
  mutate(SEQN = as.numeric(as.character(SEQN)))

data_clr <- full_join(data_cov, data_clr, by="SEQN")
dim(data_clr)

####
# design object

nhanes_design_clr <- svydesign(
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMEC4YR,
  data = data_clr,
  nest = TRUE
)

nhanes_design_clr <- subset(nhanes_design_clr, SEQN %in% seqn_sub)
nrow(nhanes_design_clr) # 9349
```

Now run model on CLR transformed data, only including the filtered taxa.

```{r}
model_clr <- list()

for(i in taxa_id_sub){
  model_clr[[i]] <- svyglm(as.formula(paste0(i, "~ quart_DR1TCAFF + RIDAGEYR_cat + RIAGENDR + RIDRETH1")), nhanes_design_clr, family=gaussian())
}

model_clr_wald <- llply(model_clr, function(x){
  regTermTest(x, "quart_DR1TCAFF")
})

# example result for RSV_genus1007
summary(model_clr[[1]])
model_clr_wald[1]

# make a results table with the wald p-values
model_clr_res <- llply(model_clr_wald, function(x){
  x$p
}) %>% 
  ldply(., .id = "taxa_id") %>% 
  rename(pval = "1")

# multiple testing correction
dim(model_clr_res)
summary(model_clr_res$pval)

model_clr_res <- model_clr_res %>% 
  mutate(
    adj_pval = p.adjust(pval, method = "bonferroni", n=length(pval))
  )

summary(model_clr_res$adj_pval)

# number of significant taxa (after bonferroni)
model_clr_res %>% 
  filter(adj_pval < 0.05)
## there are 6

# link taxa name to results table
model_clr_res <- model_clr_res %>% 
  arrange(adj_pval, pval) %>% 
  left_join(., data_taxaname, by = "taxa_id") %>% 
  select(taxa_id, taxa_name, everything(.))

# look at the significant taxa
model_clr_res %>% 
  filter(adj_pval <0.05)

# look at top 10
head(model_clr_res, 10)
```

```{r}
rm(model_clr)
```

